---
title: "Research Analysis Notebook"
---

```{r}
#label: timestamp-start

timestamp()
```


```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```

```{r}
load(
  here(
    "data",
    "data_constructed",
    "analytical_data.RData"
  )
)
```

```{r}
# checking 
names(final_data)
```

# Research Question

This project examines whether Renewable Portfolio Standards (RPS) are associated with higher levels of renewable electricity generation. Specifically, I investigate whether states with more stringent renewable energy mandates generate a larger share of their electricity from renewable sources.

```{r}
# Descriptive Statistics 1
final_data |>
  summarise(
    mean_renewable = mean(percent_renewable, na.rm = TRUE),
    sd_renewable = sd(percent_renewable, na.rm = TRUE),
    mean_rps = mean(rps_percent, na.rm = TRUE),
    sd_rps = sd(rps_percent, na.rm = TRUE)
  )

# Descriptive Statistics 2
datasummary_skim(
  final_data
)
```


# Interpreting Descriptive Statistics 
The table presents descriptive statistics for the state-year panel dataset covering 32 jurisdictions with Renewable Portfolio Standards (RPS) between 2000 and 2023 (N = 768 observations). The average share of electricity generated from renewable sources was 42.2%, with substantial variation across observations (SD = 38.0). Renewable generation shares ranged from less than 1% to 100%, indicating considerable heterogeneity in renewable energy adoption across states and over time. The median renewable share (26.3%) is notably lower than the mean, suggesting a right-skewed distribution in which a subset of states achieved exceptionally high levels of renewable generation.

The key independent variable, Renewable Portfolio Standards, also exhibits substantial variation. Across all state-year observations, the average RPS requirement was 9.1%, with values ranging from 0% to 63%. The median RPS requirement of 5.4% indicates that many observations correspond to relatively modest policy mandates, while a smaller number of states adopted more ambitious renewable energy targets. This variation is important for identifying the relationship between policy stringency and renewable energy outcomes.

The dataset contains a balanced panel of 32 jurisdictions observed annually over a 24-year period, with each state contributing an equal number of observations. Renewable energy generation averaged approximately 124,000 billion Btu, while total energy generation averaged roughly 1.29 million billion Btu. Both variables exhibit substantial dispersion, reflecting large differences in the scale of energy systems across states. The wide range of renewable generation shares, together with meaningful variation in RPS requirements, provides an appropriate empirical foundation for examining whether stronger renewable energy mandates are associated with greater renewable electricity production.

The descriptive statistics also highlight an important characteristic of the sample. Although all jurisdictions included in the analysis adopted some form of Renewable Portfolio Standard during the study period, the stringency and timing of these policies varied considerably. Consequently, the analysis primarily examines variation among RPS-adopting states rather than differences between adopting and non-adopting states. The results should therefore be interpreted as evidence regarding the effectiveness of stronger versus weaker renewable energy mandates within the population of states that implemented RPS policies.

```{r}
# PRS and renewable generation 
library(ggplot2)

ggplot(
  final_data,
  aes(
    x = rps_percent,
    y = percent_renewable
  )
) +
  geom_point(alpha = 0.6) +
  geom_smooth(
    method = "lm",
    se = TRUE
  ) +
  labs(
    x = "Renewable Portfolio Standard (%)",
    y = "Renewable Generation Share (%)",
    title = "Relationship between RPS and Renewable Generation"
  ) +
  theme_minimal()
```

# Interpreting the regression plot

The regression figure displays the relationship between Renewable Portfolio Standards (RPS) and renewable electricity generation. The fitted regression line indicates a positive relationship, suggesting that states with stronger renewable energy mandates tend to generate a larger share of their electricity from renewable sources. As RPS requirements increase, renewable generation generally increases as well.At the same time, the figure reveals substantial variation around the trend line. States with similar RPS requirements often exhibit very different levels of renewable generation, indicating that factors beyond policy mandates influence renewable energy outcomes. Natural resource availability, geographic conditions, electricity market structures, historical energy investments, and political support for renewable energy may all contribute to these differences.A notable feature of the figure is the large concentration of observations at the upper boundary of 100% renewable generation. Approximately 18% of observations fall at this maximum value, creating a visible horizontal cluster across the top of the plot. This pattern suggests that many state-year observations achieved extremely high levels of renewable electricity generation regardless of their RPS requirement. Consequently, the relationship between RPS and renewable generation appears positive but far from deterministic.The figure also shows a substantial number of observations with RPS values close to zero. Some of these observations nevertheless exhibit relatively high renewable generation shares, suggesting that renewable energy production may result from favorable resource endowments or historical energy systems even in the absence of strong renewable portfolio mandates.

```{r}
# Basic OLS model; Bivariate Relationship
model1 <- lm(
  percent_renewable ~ rps_percent,
  data = final_data
)

summary(model1)
```


# Interpreting result of OLS Model: Model 1
To examine the relationship between Renewable Portfolio Standards (RPS) and renewable electricity generation, an ordinary least squares (OLS) regression model was estimated with renewable electricity generation share as the dependent variable and RPS requirements as the independent variable.
The results indicate a positive and statistically significant relationship between RPS requirements and renewable electricity generation (β = 0.823, p < .001). Specifically, a one-percentage-point increase in a state's RPS requirement is associated with an estimated 0.82 percentage-point increase in the share of electricity generated from renewable sources. This finding suggests that states with stronger renewable energy mandates tend to produce larger shares of renewable electricity.
The model is statistically significant overall (F(1,766) = 48.48, p < .001), indicating that RPS requirements are systematically associated with renewable electricity generation. However, the explanatory power of the model remains relatively modest. The R² value of 0.060 indicates that RPS requirements explain approximately 6% of the variation in renewable electricity generation across state-year observations. While this suggests that RPS policies are an important predictor of renewable energy outcomes, most of the variation remains attributable to factors not included in the model.
Overall, the findings provide preliminary evidence that stronger Renewable Portfolio Standards are associated with higher levels of renewable electricity generation. Nevertheless, renewable energy production is likely influenced by a broader set of factors, including natural resource availability, geographic conditions, electricity market structures, technological development, and state-specific policy environments. Subsequent fixed-effects models can help determine whether this positive relationship persists after accounting for differences across states and over time.

```{r}
model2 <- lm(
  percent_renewable ~ rps_percent + factor(state),
  data = final_data
)

summary(model2)

```


# Interpreting State Fixed Effect Model: Model 2
To account for unobserved differences across states, a state fixed-effects model was estimated. This specification controls for time-invariant state characteristics that may influence renewable electricity generation, such as geography, renewable resource availability, historical energy infrastructure, political culture, and long-standing energy policies. By focusing on within-state changes over time, the model provides a more rigorous assessment of the relationship between Renewable Portfolio Standards (RPS) and renewable electricity generation.
The results indicate a positive and highly statistically significant relationship between RPS requirements and renewable electricity generation (β = 0.577, p < .001). Specifically, a one-percentage-point increase in a state's RPS requirement is associated with an estimated 0.58 percentage-point increase in the share of electricity generated from renewable sources. Substantively, a 10-percentage-point increase in a state's RPS target is associated with approximately a 5.8 percentage-point increase in renewable electricity generation.
Compared to the baseline OLS model, the estimated effect remains positive and statistically significant, although the magnitude is somewhat smaller. This finding suggests that part of the association observed in the OLS model reflects persistent differences between states. However, even after controlling for these stable state characteristics, increases in RPS requirements remain associated with higher levels of renewable electricity generation. This result provides stronger evidence that renewable portfolio standards contribute to renewable energy development rather than merely reflecting differences between states that are already more supportive of renewable energy.
Overall, the fixed-effects results strengthen confidence in the relationship between RPS policies and renewable electricity generation by demonstrating that the positive association persists when comparing changes within states over time rather than relying solely on cross-state differences.


```{r}
# model 3: state and year fixed effect
library(fixest)

model3 <- feols(
  percent_renewable ~ rps_percent |
    state + year,
  data = final_data
)

summary(model3)
```

# Interpreting state and year fixed effect: Model 3
To further examine the relationship between Renewable Portfolio Standards (RPS) and renewable electricity generation, a state and year fixed-effects model was estimated. This specification controls for both time-invariant differences across states and factors that change over time but affect all states similarly. State fixed effects account for characteristics such as geography, renewable resource availability, political culture, and historical energy infrastructure, while year fixed effects capture national trends, technological developments, and broader economic conditions that may influence renewable energy generation.
The results indicate a positive and highly statistically significant relationship between RPS requirements and renewable electricity generation (β = 0.448, p < .001). Specifically, a one-percentage-point increase in a state's RPS requirement is associated with an estimated 0.45 percentage-point increase in the share of electricity generated from renewable sources. Substantively, a 10-percentage-point increase in RPS requirements is associated with approximately a 4.5 percentage-point increase in renewable electricity generation.
Compared to the baseline OLS model (β = 0.823) and the state fixed-effects model (β = 0.577), the estimated effect decreases in magnitude but remains highly significant. This pattern suggests that some of the association observed in simpler models reflects persistent differences between states and broader temporal trends. Nevertheless, the continued significance of the coefficient after accounting for both state and year effects provides stronger evidence that Renewable Portfolio Standards contribute to increases in renewable electricity generation.
Overall, the state and year fixed-effects model provides the most rigorous test of the relationship examined in this study. The findings indicate that increases in RPS requirements are associated with higher levels of renewable electricity generation within states over time, even after accounting for state-specific characteristics and nationwide trends affecting renewable energy development.


```{r}
library(modelsummary)

modelsummary(
  list(
    "OLS" = model1,
    "State FE" = model2,
    "State + Year FE" = model4
  ),
  statistic = "({std.error})",
  stars = c("*" = .05, "**" = .01, "***" = .001),
  gof_omit = "AIC|BIC|Log.Lik"
)
```

# Comparing three models
The above table reprenst three models examining the relationship between Renewable Portfolio Standards (RPS) and the share of electricity generated from renewable sources. In the simplest bivariate OLS model (Model 1), RPS requirements are positively and statistically significantly associated with renewable generation (β = 0.823, p < .001). This suggests that a one-percentage-point increase in a state's RPS requirement is associated with approximately a 0.82 percentage-point increase in the share of renewable electricity generation. However, the model explains only about 6% of the variation in renewable generation (R² = .060), indicating that many other factors influence renewable energy outcomes.

Model 2 introduces state fixed effects, controlling for time-invariant characteristics of states such as geography, resource endowments, political culture, and historical energy infrastructure. The estimated effect of RPS remains positive and statistically significant (β = 0.577, p < .001), although the coefficient decreases in magnitude. This suggests that part of the relationship observed in Model 1 reflects differences between states, but even after accounting for these differences, states tend to generate a larger share of renewable electricity when they adopt stronger RPS requirements. The substantial increase in model fit (R² = .961) indicates that persistent differences among states explain a large share of the variation in renewable generation.

The most rigorous specification is Model 3, which includes both state and year fixed effects. In this model, the RPS coefficient remains positive and statistically significant (β = 0.448, p < .001). This coefficient can be interpreted as a within-state effect: when a state's RPS requirement increases by one percentage point, the share of renewable electricity generation is associated with an increase of approximately 0.45 percentage points, holding constant both state-specific characteristics and nationwide changes affecting all states in a given year. Although the coefficient is smaller than in previous models, its continued significance provides stronger evidence that RPS policies are associated with higher renewable electricity generation rather than merely reflecting differences across states or broader temporal trends.

As more controls are introduced, the estimated effect of RPS becomes smaller but remains statistically significant. It suggests that part of the raw relationship is explained by underlying differences between states, yet a meaningful positive association persists even after accounting for state-specific and year-specific factors. Therefore, the findings provide evidence that stronger Renewable Portfolio Standards are associated with greater renewable electricity generation, although the magnitude of the relationship is more modest once unobserved heterogeneity is taken into account.

```{r}
#label: timestamp-end

timestamp()
```