---
title: "Organize Data Notebook"
---
#| label: load-data
#| message: false
#| warning: false

library(readxl)
library(tidyverse)
library(here)
library(modelsummary)
library(gt)
library(here)
library(fixest)

```{r}
#label: timestamp-start

timestamp()
```

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```

This quarto doc is used to organize the data.

# Load the Data

```{r}
# Load the data
RPS <- read_excel(here("data", "data_raw", "RPS.xlsx"))
Renewable_Generation1 <- read_excel(here("data", "data_raw", "Total Renewable energy production.xlsx"))
Renewable_Generation2 <- read_excel(here("data", "data_raw", "Total Renewable energy production.xlsx"), sheet = 2)

```


```{r}
# checking Basic structure
glimpse(RPS)
glimpse(Renewable_Generation1)
glimpse(Renewable_Generation2)
```

```{r}
# checking Column names
names(RPS)
names(Renewable_Generation1)
names(Renewable_Generation2)
names(energy_data)
```

```{r}
# checking First few rows
head(RPS)
head(Renewable_Generation1)
head(Renewable_Generation2)
```

```{r}
# clean rps
RPS <- read_excel(here("data", "data_raw", "RPS.xlsx"), skip = 24)
RPS <- RPS |> 
  rename(state = State, rps_tier = `RPS Tier or Carve Out`) |>
  #  Total RPS
  filter(str_detect(rps_tier, "^Total RPS")) |>
  select(state, starts_with("20")) |>
  pivot_longer(cols = starts_with("20"), names_to = "year", 
               values_to = "rps_total") |>
  mutate(year = as.numeric(year))

# clean 2: reverting states names from abbreviation to full name 
RPS <- RPS |>
  mutate(
    state = recode(state,
      !!!setNames(state.name, state.abb),
      "DC" = "District of Columbia"
    )
  )

```


```{r}
# cleaning renewable energy generation  
# fixing column names

# Extracting the second row as column names
names(Renewable_Generation1) <- as.character(
  Renewable_Generation1[2, ]
)

# Removing the first two rows
Renewable_Generation1 <- Renewable_Generation1[-c(1,2), ]

# 2
names(Renewable_Generation2) <- as.character(
  Renewable_Generation2[2, ]
)

# Remove the first two rows for sheet 2
Renewable_Generation2 <- Renewable_Generation2[-c(1,2), ]


# keep ultimate years from 2000 to 2023
Renewable_Generation1 <- Renewable_Generation1 |>
  select(State, any_of(as.character(2000:2023)))

Renewable_Generation2 <- Renewable_Generation2 |>
  select(State, any_of(as.character(2000:2023)))

```

```{r}
# merging RPS and renewable generation 
energy_data <- Renewable_Generation1 |>
  left_join(
    Renewable_Generation2,
    by = c("State", "year")
  ) |>
  mutate(
    percent_renewable = 100 * renewable_generation / total_generation,
    state = case_when(
      State == "DC" ~ "District of Columbia",
      TRUE ~ state.name[match(State, state.abb)]
    )
  ) |>
  select(
    state,
    year,
    renewable_generation,
    total_generation,
    percent_renewable
  )

# check
RPS_states <- unique(RPS$state)

energy_data <- energy_data |>
  filter(state %in% RPS_states)


# check
length(unique(energy_data$state))

setdiff(unique(RPS$state), unique(energy_data$state))

# final dataset 
final_data <- energy_data |>
  left_join(RPS, by = c("state", "year"))

# check final dataset
glimpse(final_data)

sum(is.na(final_data$rps_total))

setdiff(unique(RPS$state), unique(final_data$state))

# calculate RPS as percentage 
final_data <- final_data |>
  mutate(
    rps_percent = rps_total * 100
  )
```



```{r}
# save data
save(
  final_data,
  file = here(
    "data",
    "data_constructed",
    "analytical_data.RData"
  )
)
```


```{r}
# Descriptive Statistics
library(dplyr)

final_data |>
  summarise(
    mean_renewable = mean(percent_renewable, na.rm = TRUE),
    sd_renewable = sd(percent_renewable, na.rm = TRUE),
    mean_rps = mean(rps_percent, na.rm = TRUE),
    sd_rps = sd(rps_percent, na.rm = TRUE)
  )
```


```{r}
# Descriptive statistics 
datasummary_skim(
  final_data
)

# 
library(modelsummary)

datasummary_skim(
  final_data |> select(percent_renewable, rps_percent)
)
```

```{r}
# Descriptive Statistics 
final_data |>
  summarise(
    n = n(),
    mean_renewable = mean(percent_renewable, na.rm = TRUE),
    sd_renewable = sd(percent_renewable, na.rm = TRUE),
    min_renewable = min(percent_renewable, na.rm = TRUE),
    max_renewable = max(percent_renewable, na.rm = TRUE),

    mean_rps = mean(rps_percent, na.rm = TRUE),
    sd_rps = sd(rps_percent, na.rm = TRUE),
    min_rps = min(rps_percent, na.rm = TRUE),
    max_rps = max(rps_percent, na.rm = TRUE)
  )
```

# Interpretation 
Table 1 presents descriptive statistics for the state-year panel dataset covering 32 jurisdictions with Renewable Portfolio Standards (RPS) between 2000 and 2023 (N = 768 observations). The average share of electricity generated from renewable sources was 42.2%, with substantial variation across observations (SD = 38.0). Renewable generation shares ranged from less than 1% to 100%, indicating considerable heterogeneity in renewable energy adoption across states and over time. The median renewable share (26.3%) is notably lower than the mean, suggesting a right-skewed distribution in which a subset of states achieved exceptionally high levels of renewable generation.

The key independent variable, Renewable Portfolio Standards, also exhibits substantial variation. Across all state-year observations, the average RPS requirement was 9.1%, with values ranging from 0% to 63%. The median RPS requirement of 5.4% indicates that many observations correspond to relatively modest policy mandates, while a smaller number of states adopted more ambitious renewable energy targets. This variation is important for identifying the relationship between policy stringency and renewable energy outcomes.

The dataset contains a balanced panel of 32 jurisdictions observed annually over a 24-year period, with each state contributing an equal number of observations. Renewable energy generation averaged approximately 124,000 billion Btu, while total energy generation averaged roughly 1.29 million billion Btu. Both variables exhibit substantial dispersion, reflecting large differences in the scale of energy systems across states. The wide range of renewable generation shares, together with meaningful variation in RPS requirements, provides an appropriate empirical foundation for examining whether stronger renewable energy mandates are associated with greater renewable electricity production.

The descriptive statistics also highlight an important characteristic of the sample. Although all jurisdictions included in the analysis adopted some form of Renewable Portfolio Standard during the study period, the stringency and timing of these policies varied considerably. Consequently, the analysis primarily examines variation among RPS-adopting states rather than differences between adopting and non-adopting states. The results should therefore be interpreted as evidence regarding the effectiveness of stronger versus weaker renewable energy mandates within the population of states that implemented RPS policies.

```{r}
# PRS and renewable generation 
library(ggplot2)

ggplot(
  final_data,
  aes(
    x = rps_percent,
    y = percent_renewable
  )
) +
  geom_point(alpha = 0.6) +
  geom_smooth(
    method = "lm",
    se = TRUE
  ) +
  labs(
    x = "Renewable Portfolio Standard (%)",
    y = "Renewable Generation Share (%)",
    title = "Relationship between RPS and Renewable Generation"
  ) +
  theme_minimal()
```

# Interpretation 
Figure 1 displays the relationship between Renewable Portfolio Standards (RPS) and renewable electricity generation. The fitted regression line indicates a positive relationship, suggesting that states with stronger renewable energy mandates tend to generate a larger share of their electricity from renewable sources. As RPS requirements increase, renewable generation generally increases as well.
At the same time, the figure reveals substantial variation around the trend line. States with similar RPS requirements often exhibit very different levels of renewable generation, indicating that factors beyond policy mandates influence renewable energy outcomes. Natural resource availability, geographic conditions, electricity market structures, historical energy investments, and political support for renewable energy may all contribute to these differences.
A notable feature of the figure is the large concentration of observations at the upper boundary of 100% renewable generation. Approximately 18% of observations fall at this maximum value, creating a visible horizontal cluster across the top of the plot. This pattern suggests that many state-year observations achieved extremely high levels of renewable electricity generation regardless of their RPS requirement. Consequently, the relationship between RPS and renewable generation appears positive but far from deterministic.
The figure also shows a substantial number of observations with RPS values close to zero. Some of these observations nevertheless exhibit relatively high renewable generation shares, suggesting that renewable energy production may result from favorable resource endowments or historical energy systems even in the absence of strong renewable portfolio mandates.


```{r}
# Figure - boxplot 
ggplot(
  final_data,
  aes(
    x = factor(year),
    y = percent_renewable
  )
) +
  geom_boxplot() +
  labs(
    x = "Year",
    y = "Renewable Generation Share (%)",
    title = "Distribution of Renewable Generation Share by Year"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
# Interpretation 
Figure 2 presents the distribution of renewable electricity generation shares across states from 2000 to 2023. Several important patterns emerge.
First, the median renewable generation share increases over time, rising from approximately 20% in the early 2000s to roughly 35% by 2023. This suggests that renewable energy has become a progressively larger component of state electricity portfolios during the study period.
Second, the upper portion of the distribution shifts upward substantially over time. By the mid-2010s and especially after 2020, the upper quartiles approach or reach 100%, indicating that a growing number of state-year observations achieved very high levels of renewable electricity generation. This pattern is consistent with the expansion of renewable energy technologies, increasing policy support, declining technology costs, and broader decarbonization efforts across many states.
Third, considerable variation remains across states throughout the entire period. Even in recent years, some states exhibit relatively low renewable generation shares while others approach complete reliance on renewable electricity sources. The persistence of wide interquartile ranges indicates substantial heterogeneity in state energy systems, resource endowments, policy environments, and infrastructure capacities.
Finally, the figure suggests a gradual upward trend rather than an abrupt change. Renewable generation appears to increase steadily over the study period, which aligns with expectations that energy transitions occur incrementally due to the long planning horizons and capital investments required for electricity infrastructure.
```{r}
ggplot(
  final_data,
  aes(
    x = year,
    y = percent_renewable,
    group = state
  )
) +
  geom_line(alpha = 0.35) +
  labs(
    x = "Year",
    y = "Renewable Generation Share (%)",
    title = "State-Level Renewable Generation Trends"
  ) +
  theme_minimal()
```
# Interpretation 

Figure 3 presents state-level trends in renewable electricity generation shares between 2000 and 2023. Each line represents a state's renewable generation share over time. The figure reveals substantial heterogeneity in renewable energy trajectories across states.
Several states exhibit consistently high renewable generation shares throughout the study period, maintaining levels above 75% and in some cases approaching 100%. Other states begin with relatively low renewable generation shares but experience gradual increases over time. A third group of states remains at comparatively low levels throughout the period, indicating limited growth in renewable electricity production.
Overall, the figure suggests that renewable electricity generation has generally increased across many states over the last two decades. However, the pace and magnitude of change vary considerably. While some states experienced substantial growth in renewable generation, others showed only modest increases. This variation highlights the importance of state-specific factors, including natural resource endowments, energy infrastructure, electricity market structures, policy environments, and economic conditions.
One particularly notable pattern is the presence of several states that experience sharp increases in renewable generation during the study period. These abrupt changes may reflect major investments in renewable energy infrastructure, changes in energy reporting practices, or significant policy shifts. In contrast, many states exhibit more gradual upward trends, consistent with the long-term nature of energy system transitions.
```{r}
# Basic OLS model; Bivariate Relationship
model1 <- lm(
  percent_renewable ~ rps_percent,
  data = final_data
)

summary(model1)
```
# Interpreting result of OLS Model
To examine the relationship between Renewable Portfolio Standards (RPS) and renewable electricity generation, an ordinary least squares (OLS) regression model was estimated with renewable electricity generation share as the dependent variable and RPS requirements as the independent variable.
The results indicate a positive and statistically significant relationship between RPS requirements and renewable electricity generation (β = 0.823, p < .001). Specifically, a one-percentage-point increase in a state's RPS requirement is associated with an estimated 0.82 percentage-point increase in the share of electricity generated from renewable sources. This finding suggests that states with stronger renewable energy mandates tend to produce larger shares of renewable electricity.
The model is statistically significant overall (F(1,766) = 48.48, p < .001), indicating that RPS requirements are systematically associated with renewable electricity generation. However, the explanatory power of the model remains relatively modest. The R² value of 0.060 indicates that RPS requirements explain approximately 6% of the variation in renewable electricity generation across state-year observations. While this suggests that RPS policies are an important predictor of renewable energy outcomes, most of the variation remains attributable to factors not included in the model.
Overall, the findings provide preliminary evidence that stronger Renewable Portfolio Standards are associated with higher levels of renewable electricity generation. Nevertheless, renewable energy production is likely influenced by a broader set of factors, including natural resource availability, geographic conditions, electricity market structures, technological development, and state-specific policy environments. Subsequent fixed-effects models can help determine whether this positive relationship persists after accounting for differences across states and over time.

```{r}
# model1 summary 
library(modelsummary)

modelsummary(
  model1,
  statistic = "({std.error})",
  stars = TRUE
)
```


```{r}
# State Fixed Effect Model 
model2 <- lm(
  percent_renewable ~ rps_percent + factor(state),
  data = final_data
)

summary(model2)

#
library(fixest)

model2 <- feols(
  percent_renewable ~ rps_percent | state,
  data = final_data
)
```

# Interpretation
To account for unobserved differences across states, a state fixed-effects model was estimated. This specification controls for time-invariant state characteristics that may influence renewable electricity generation, such as geography, renewable resource availability, historical energy infrastructure, political culture, and long-standing energy policies. By focusing on within-state changes over time, the model provides a more rigorous assessment of the relationship between Renewable Portfolio Standards (RPS) and renewable electricity generation.
The results indicate a positive and highly statistically significant relationship between RPS requirements and renewable electricity generation (β = 0.577, p < .001). Specifically, a one-percentage-point increase in a state's RPS requirement is associated with an estimated 0.58 percentage-point increase in the share of electricity generated from renewable sources. Substantively, a 10-percentage-point increase in a state's RPS target is associated with approximately a 5.8 percentage-point increase in renewable electricity generation.
Compared to the baseline OLS model, the estimated effect remains positive and statistically significant, although the magnitude is somewhat smaller. This finding suggests that part of the association observed in the OLS model reflects persistent differences between states. However, even after controlling for these stable state characteristics, increases in RPS requirements remain associated with higher levels of renewable electricity generation. This result provides stronger evidence that renewable portfolio standards contribute to renewable energy development rather than merely reflecting differences between states that are already more supportive of renewable energy.
Overall, the fixed-effects results strengthen confidence in the relationship between RPS policies and renewable electricity generation by demonstrating that the positive association persists when comparing changes within states over time rather than relying solely on cross-state differences.

```{r}
# model 3: state and year fixed effect
library(fixest)

model3 <- feols(
  percent_renewable ~ rps_percent |
    state + year,
  data = final_data
)

summary(model3)
```


# Interpretation 
To further examine the relationship between Renewable Portfolio Standards (RPS) and renewable electricity generation, a state and year fixed-effects model was estimated. This specification controls for both time-invariant differences across states and factors that change over time but affect all states similarly. State fixed effects account for characteristics such as geography, renewable resource availability, political culture, and historical energy infrastructure, while year fixed effects capture national trends, technological developments, and broader economic conditions that may influence renewable energy generation.
The results indicate a positive and highly statistically significant relationship between RPS requirements and renewable electricity generation (β = 0.448, p < .001). Specifically, a one-percentage-point increase in a state's RPS requirement is associated with an estimated 0.45 percentage-point increase in the share of electricity generated from renewable sources. Substantively, a 10-percentage-point increase in RPS requirements is associated with approximately a 4.5 percentage-point increase in renewable electricity generation.
Compared to the baseline OLS model (β = 0.823) and the state fixed-effects model (β = 0.577), the estimated effect decreases in magnitude but remains highly significant. This pattern suggests that some of the association observed in simpler models reflects persistent differences between states and broader temporal trends. Nevertheless, the continued significance of the coefficient after accounting for both state and year effects provides stronger evidence that Renewable Portfolio Standards contribute to increases in renewable electricity generation.
Overall, the state and year fixed-effects model provides the most rigorous test of the relationship examined in this study. The findings indicate that increases in RPS requirements are associated with higher levels of renewable electricity generation within states over time, even after accounting for state-specific characteristics and nationwide trends affecting renewable energy development.


```{r}
library(modelsummary)

modelsummary(
  list(
    "OLS" = model1,
    "State FE" = model2,
    "OLS + Year Trend" = model3,
    "State + Year FE" = model4
  ),
  statistic = "({std.error})",
  stars = c("*" = .05, "**" = .01, "***" = .001),
  gof_omit = "AIC|BIC|Log.Lik"
)
```

# Interpretation 
Table X presents four models examining the relationship between Renewable Portfolio Standards (RPS) and the share of electricity generated from renewable sources. In the simplest bivariate OLS model (Model 1), RPS requirements are positively and statistically significantly associated with renewable generation (β = 0.823, p < .001). This suggests that a one-percentage-point increase in a state's RPS requirement is associated with approximately a 0.82 percentage-point increase in the share of renewable electricity generation. However, the model explains only about 6% of the variation in renewable generation (R² = .060), indicating that many other factors influence renewable energy outcomes.
Model 2 introduces state fixed effects, controlling for time-invariant characteristics of states such as geography, resource endowments, political culture, and historical energy infrastructure. The estimated effect of RPS remains positive and statistically significant (β = 0.577, p < .001), although the coefficient decreases in magnitude. This suggests that part of the relationship observed in Model 1 reflects differences between states, but even after accounting for these differences, states tend to generate a larger share of renewable electricity when they adopt stronger RPS requirements. The substantial increase in model fit (R² = .961) indicates that persistent differences among states explain a large share of the variation in renewable generation.
Model 3 adds a linear time trend instead of state fixed effects. The coefficient for RPS remains positive and significant (β = 0.897, p < .001), while the year variable is not statistically significant (β = -0.207, p > .05). This suggests that a simple national time trend does not explain much of the observed relationship between RPS policies and renewable generation. The model's explanatory power remains nearly identical to Model 1 (R² = .060), indicating that year alone contributes little additional information.
The most rigorous specification is Model 4, which includes both state and year fixed effects. In this model, the RPS coefficient remains positive and statistically significant (β = 0.448, p < .001). This coefficient can be interpreted as a within-state effect: when a state's RPS requirement increases by one percentage point, the share of renewable electricity generation is associated with an increase of approximately 0.45 percentage points, holding constant both state-specific characteristics and nationwide changes affecting all states in a given year. Although the coefficient is smaller than in previous models, its continued significance provides stronger evidence that RPS policies are associated with higher renewable electricity generation rather than merely reflecting differences across states or broader temporal trends.

As more controls are introduced, the estimated effect of RPS becomes smaller but remains statistically significant. This is generally what researchers hope to see. It suggests that part of the raw relationship is explained by underlying differences between states, yet a meaningful positive association persists even after accounting for state-specific and year-specific factors. Therefore, the findings provide evidence that stronger Renewable Portfolio Standards are associated with greater renewable electricity generation, although the magnitude of the relationship is more modest once unobserved heterogeneity is taken into account.

The two-way fixed effects model provides the most credible estimate, suggesting that a one-percentage-point increase in a state's RPS requirement is associated with a 0.45 percentage-point increase in renewable electricity generation after controlling for both state and year fixed effects. This indicates that RPS policies are positively associated with renewable energy development, although the effect is smaller than implied by simple cross-sectional comparisons.
```{r}
#label: timestamp-end

timestamp()
```