---
# basic information
title: "State Renewable Energy Policies and Renewable Electricity Generation: An Analysis of Renewable Portfolio Standards in 2008–2009"
keywords: [Renewable Portfolio Standards, renewable energy production, US States]

# information used by aog-article format
shorttitle: "Paper"
shortauthors: Samane Afshari

## bibliographic information is in the _quarto.yml for the project ##

# I use the abstact-section extension to write abstracts as a normal section
# of the document. If you prefer to write abstracts in yaml, you can comment
# out the filter and uncomment the abstract line.
#abstract: An Abstract
filters:
  - abstract-section

# default formats
# I use a custom aog-article-pdf available as an extension, but this can be 
# replaced with the default pdf format. The submittable extenstion
# also gives me submittable-pdf and submittable-docx for journal submission
# entries, but I leave those commented out until I need them.
format:
  aog-article-pdf:
    keep-tex: false
  #pdf: 
  #  keep-tex: false
  #submittable-pdf:
  #  keep-tex: false
  #  titlepage: true
  #  endnotes: false
  #  fig-pos: "!t"
  #submittable-docx: default
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```

```{r}
load(
  here(
    "data",
    "data_constructed",
    "analytical_data.RData"
  )
)
```

# Abstract

.....

# Introduction

The transition toward renewable energy has become a central policy objective in the United States as concerns over climate change, carbon emissions, and fossil fuel dependence continue to intensify. In the absence of a unified federal leadership, state governments have increasingly emerged as major regulator in renewable energy governance through different energy-related policy instruments. 

Renewable Portfolio Standards (RPS) are among the most widely used state-level policy mechanisms because they require utilities to generate a specified share of electricity from renewable sources such as wind, solar, geothermal, and biomass. By 2008, more than half of U.S. states had adopted mandatory RPS programs, reflecting a broader shift toward decentralized energy governance and state-led climate action. 

Despite the rapid expansion of state renewable energy policies, there remains ongoing debate regarding whether these policies actually increase renewable energy production. Using PRS and Renewable Electricity Generation as the main variables, this study examines whether state-level renewable energy policies increase renewable electricity generation across U.S. states during 2008 and 2009. 

Existing literature has produced mixed findings on the effectiveness of Renewable Portfolio Standards. Early studies often relied on qualitative analyses or case studies that highlighted successful examples of renewable energy deployment but provided limited empirical evidence regarding broader policy effectiveness. More recent quantitative research has attempted to evaluate the relationship between RPS policies and renewable energy generation. For example, Carley (2009) examined whether state RPS adoption increased the share of renewable electricity generation across states and found that while RPS policies increased the total amount of renewable generation, they did not significantly increase renewable energy’s share within total electricity production.

This study builds directly on Carley’s (2009) evaluation of state renewable electricity policies, as it extends Carley’s core policy question into the 2008–2009 period, when more states had adopted RPS programs and policy designs had become more developed. This study offers a focused short-term reassessment of whether state renewable energy mandates were beginning to shift the composition of state electricity systems, not simply increase renewable generation in absolute terms. [@carley2009].


# Background
....

# Data and Methods
Independent Variable (IV): State Renewable Portfolio Standards (RPS)
The independent variable is derived from the dataset U.S. State Electricity Resource Standards: 2024 Status Update, compiled by the Lawrence Berkeley National Laboratory (LBNL). The dataset provides information on state-level Renewable Portfolio Standards (RPS) and Clean Energy Standards (CES) across U.S. states, including policy adoption dates, renewable energy targets, and compliance timelines. For this study, the dataset can be used to measure whether a state has adopted an RPS policy and the strength of the policy target over time, though the research only focus on 2008 and 2009. 


Dependent Variable (DV): Renewable Electricity Generation Share
The dependent variable is drawn from the U.S. Energy Information Administration (EIA) Annual Energy Review renewable electricity generation dataset. The dataset provides state-level information on electricity generation and the percentage of generation derived from renewable energy sources in 2008 and 2009. In this study, the dependent variable is the share of electricity generation from renewable energy sources, with particular attention to the “percent non-hydro renewable” measure, which captures renewable generation from sources such as wind, solar, geothermal, and biomass while excluding conventional hydroelectric power. This measure is appropriate because it reflects the types of renewable energy sources most directly influenced by modern RPS policies.


# Results

The table presents the results from four regression models examining the relationship between Renewable Portfolio Standards (RPS) and renewable electricity generation. The baseline OLS model indicates a positive and statistically significant relationship between RPS requirements and renewable electricity generation (β = 0.587, p < .05). Substantively, a one-percentage-point increase in a state's RPS target is associated with approximately a 0.59 percentage-point increase in the share of electricity generated from renewable sources. This finding initially suggests that states with more ambitious renewable energy policies tend to generate a larger share of their electricity from renewable sources.

However, the magnitude and statistical significance of the RPS coefficient decline substantially as more rigorous model specifications are introduced. After controlling for state fixed effects, the coefficient decreases to 0.266 and is no longer statistically significant. This result suggests that much of the positive relationship observed in the OLS model reflects persistent differences between states rather than the effects of policy changes within states. States with stronger RPS requirements may also possess favorable renewable energy resources, supportive political environments, or preexisting clean energy infrastructure that contribute to higher renewable generation levels independent of policy changes.

The addition of year fixed effects further reduces the estimated effect of RPS requirements, producing a small negative coefficient (β = -0.045) that remains statistically insignificant. Accounting for year effects is important because it controls for national trends affecting all states between 2008 and 2009, such as changes in energy markets, economic conditions, or broader developments in renewable energy technologies. Once both state-specific and year-specific factors are considered, there is no evidence that short-term changes in RPS requirements were associated with measurable changes in renewable electricity generation.

The final model adds a control for total electricity generation to account for differences in the scale of state electricity systems. The inclusion of this variable has little impact on the estimated relationship between RPS requirements and renewable electricity generation. The coefficient for RPS remains close to zero and statistically insignificant (β = -0.055, p > .05), while total electricity generation itself is not significantly associated with renewable electricity share. This suggests that differences in the overall size of state electricity systems do not explain the observed patterns.

Taken together, the results indicate that the positive association observed in the baseline OLS model is not robust to the inclusion of state and year fixed effects. While states with stronger RPS policies tend to exhibit higher levels of renewable electricity generation, the analysis does not provide evidence that changes in RPS requirements between 2008 and 2009 led to significant increases in renewable electricity generation within states during that period. One plausible explanation is that renewable energy policies require longer time horizons to influence generation outcomes, as investments in renewable infrastructure often involve planning, financing, permitting, and construction processes that extend over several years. Given the limited two-year panel and relatively small sample of state-year observations, the findings should be interpreted as evidence that short-term policy effects are difficult to detect rather than as evidence that RPS policies are ineffective.

# Conclusions



<!-- Explicitly place references -->
# References
:::{#refs}
Data source: Lawrence Berkeley National Laboratory RPS Dataset
Data source: U.S. Energy Information Administration Renewable Energy Annual Data

::: {#refs}
:::

